---
title: "Semiparametric Regression"
output: html_document
---

# Preamble

http://matt-wand.utsacademics.info/SRwRS.html

Matt Wand, 1989 Harvard, Rice, Texas

Currently at UTS.

# Why Semipara reg?

Good for summarising. 

# How to do the non-linear part?

Lots of answers. We gonna use spline based approach - sticking poly together. Specifically, we use penalised splines in session 1.

# Lab 1

## My first smoothing spline

First we specify various values for the smoothing parameters.

```{r}
library(HRW)

cols <- c("dodgerblue", "magenta", 
          "red", "green3", "darkgreen")

data(WarsawApts)
x <- WarsawApts$construction.date
y <- WarsawApts$areaPerMzloty
plot(x, y, col = "dodgerblue")
penSplFit1 <- smooth.spline(x, y, spar = 1.1)
lines(penSplFit1, col = "magenta")

penSplFit2 <- smooth.spline(x, y, spar = 0.1)
lines(penSplFit2, col = "red")

penSplFit3 <- smooth.spline(x, y, spar = 0.5)
lines(penSplFit3, col = "green3", lwd = 5)
```

Data driven smoothing parameter via Generalised Cross Validation (GCV).

```{r}
penSplGCV <- smooth.spline(x, y)
plot(x, y, col = "dodgerblue")
lines(penSplGCV, col = "red")
```

GCV is also implemented in `mgcv`. The `gam()` function is much more powerful than `smooth.spline()`.

```{r}
library(mgcv)
penSplFitGCVmgcv <- gam(y~ s(x, k = 50))
plot(penSplFitGCVmgcv)
```

```{r}
shiftAmt <- mean(fitted(penSplFitGCVmgcv))
plot(penSplFitGCVmgcv, shift = shiftAmt,
     ylim = range(y))
points(x, y, col = "dodgerblue")
```


## Choosing the number of spline basis functions

Initially use the default number of basis functions (8) but based on the result for k-index this seems too low for the data.

When k is increased to 30 we get a k-index value > 1 which suggests adequate number of basis functions. The number of spline basis functions is k-2 = 28.

```{r}
fitgamDflt <- gam(y ~ s(x))
gam.check(fitgamDflt)

fitBiggerBasis <- gam(y ~ s(x, k = 30))
# gives residuals etc
gam.check(fitBiggerBasis)
```

## Degrees of freedom

The effective degrees of freedom equals 14 suggesting that the fit has about the same flexibility as a 13th degree poly. EDF is scale free - does not depend on units for x and y.

```{r}
fitBiggerBasis <- gam(y ~ s(x, k = 30))
summary(fitBiggerBasis)$edf
```

```{r}
par(mfrow = c(2, 2))
fitLowEDF <- smooth.spline(x, y, df = 2.5)
plot(x, y, col = cols[1], main = "EDF = 2.5")
lines(fitLowEDF, col = cols[5])

fitHighEDF <- smooth.spline(x, y, df = 50)
plot(x, y, col = cols[1], main = "EDF = 50")
lines(fitHighEDF, col = cols[5])

fitGCV <- smooth.spline(x, y, df = 14.087)
plot(x, y, col = cols[1], main = "EDF = 14.087")
lines(fitGCV, col = cols[5])

```


## Getting fancy vis.

Here is a publication quality plot.

```{r}
library(mgcv) ; 
fitGCVmgcv <- gam(y ~ s(x,k = 30))
par(mfrow = c(1,1))
plot(x,y,type = "n",bty = "l",
 xlab = "construction date (years)",
 ylab = "area (square metres) per million zloty",
 cex.lab = 1.5,cex.axis = 1.5)
xg <- seq(min(x),max(x),length = 1001)
fg <- predict(fitGCVmgcv,newdata
              = data.frame(x = xg),
              se.fit = TRUE)
lowg <- fg$fit - 2*fg$se.fit
uppg <- fg$fit + 2*fg$se.fit
polygon(c(xg,rev(xg)),c(lowg,rev(uppg)),
        col = "palegreen",border = FALSE)
lines(xg,fg$fit,col = "darkgreen")
points(x,y,col = "dodgerblue")
```

## Penalized spline underpinnings

```{r}
source("./SRwRSscripts/penSplinesDetails.R")
```



