---
title: "Semiparametric Regression"
output: html_document
---

# Lab 2

This uses mixed models allows more complex models to be fit. 

\begin{aligned}
y = X \beta + Z u  + e \\
\end{aligned}

Frequentist approach uses ML (for beta) or REML (for G, R) and best prediction (u). We trick mixed model into doing penalised spline smoothing by.

\begin{aligned}
y_i = \beta_0 + \beta_1 x_i + \sum_{k=1}^K u_k z_k(x)  + e_i \\
\end{aligned}

## Lab

Data gives distance from pituitary to pterygomaxillary fissure versus age.

Note $\lambda = \frac{\sigma_e^2}{\sigma_u^2}

\begin{aligned}
dist_{ij} = \beta_0 + \beta_1 age_{ij} + \beta_2 sex_{ij} + U_i + e_ij \\
\end{aligned}

The plot shows the fitted line for each child. According to the model the lines are parallel with each child having their own intercept.


```{r}
library(nlme)
plot(Orthodont)
fit <- lme(fixed = distance ~ age + Sex, random = ~1, data = Orthodont)
summary(fit)
plot(augPred(fit), grid = T)
```


## Warsaw apartments

```{r}
library(HRW)
cols <- c("dodgerblue", "magenta", 
          "red", "green3", "darkgreen")

data(WarsawApts)
x <- WarsawApts$construction.date
y <- WarsawApts$areaPerMzloty

par(mfrow = c(1, 1))
plot(x, y, xlab = "Construct date", ylab = "area per mill zloty",
     col = "dodgerblue")



```

### Tricking lme

Trick lme into fitting a penalized spline. First vis of the columns in Z: spline functions of the predictor. These are known as canonical O'Sullivan splines, see section 4 of Wand M.P 2008 *On semiparametric regression with O'Sullivan splines*.

```{r}
# trick lme into fitting a penalized spline
X <- cbind(1, x)
numIntKnots <- 25
intKnots <- quantile(unique(x), 
                     seq(0, 1, length = numIntKnots + 2)[-c(1, numIntKnots + 2)])
Z <- ZOSull(x, intKnots = intKnots, range.x = range(x))

head(Z)
plot(0, type= "n",
     xlim = range(x),
     ylim = range(Z))
for(k in 1:ncol(Z)) points(x, Z[, k], col = k)
```



```{r}
library(nlme)
dummyGp <- rep(1, length(x))
dataGp <- groupedData(y ~ x | dummyGp, data = data.frame(x, y))

penSplFitLMM <- lme(y ~ -1 + X, random = pdIdent( ~ -1 + Z), data = dataGp)

```

Extract fixed and RE.

```{r}
betaHat <- as.vector(penSplFitLMM$coefficients$fixed)
uHat <- as.vector(penSplFitLMM$coefficients$random[[1]])
print(betaHat)
print(uHat)
```

Set up grid then get the gridwise fixed effects and random effects design mats.

```{r}
ng <- 1001; xg <- seq(min(x), max(x), length = ng)
Xg <- cbind(1, xg)
Zg <- ZOSull(xg, intKnots = intKnots, range.x = range(x))
# spline
fhatg <- as.vector(Xg %*% betaHat + Zg %*% uHat)

plot(x, y, col = "dodgerblue",
     xlab = "const date",
     ylab = "area per mill")
lines(xg, fhatg, lwd = 2, col = "darkgreen")
```

### Add CI

```{r}
sigmaEpsHat <- penSplFitLMM$sigma
sigmauHat <- sqrt(as.numeric(VarCorr(penSplFitLMM)[1,1]))
lambdaHat <- sigmaEpsHat^2/sigmauHat^2
print(lambdaHat)
```

The justification of computing the conf bands as follows can be found in Ruppert D. 2003 Semiparametric regression, section 6.4.

```{r}
CTC <- crossprod(cbind(X,Z)) 
D <- diag(c(0,0,rep(1,ncol(Z))))
fullCovMatInv <- CTC + lambdaHat*D
Cg <- cbind(Xg,Zg)
sdg <- sigmaEpsHat*sqrt(
        diag(Cg%*%solve(fullCovMatInv,t(Cg))))
lowg <- fhatg - qnorm(0.975)*sdg
uppg <- fhatg + qnorm(0.975)*sdg
```

Plot

```{r}
plot(x, y, type = "n", 
     xlab = "const date",
     ylab = "area per mill")
polygon(c(xg, rev(xg)), c(lowg, rev(uppg)), col = "palegreen",
        border = F)
lines(xg, fhatg, lwd = 2, col = "darkgreen")
points(x, y, col = "dodgerblue")
```

## Use gam()

```{r}
library(mgcv)
penSplFitREMLmgcv <- gam(y~ s(x, k = 25), method = "REML")
shiftAmt <- mean(fitted(penSplFitREMLmgcv))
plot(penSplFitREMLmgcv, 
     shift = shiftAmt,
     ylim = range(y), shade = T,
     shade.col = "palegreen")

points(x, y, col = "dodgerblue")
```

















