---
title: "Semiparametric Regression"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Lab 3

You need to use at least Version 1.2.1335 of rstudio and version R-3.5.2 or later.


## Bayesian Inference

\begin{aligned}
y | \beta, u, \sigma_e^2 \sim N(X\beta + Zu, \sigma_e^2 I)
\end{aligned}

Representation for model can be done via dag.

Need to standardise.

Intractable integrals.

So, full conditionals and gibbs.

\begin{aligned}
y | \sigma^2, y \\
\sigma^2 | \mu, y
\end{aligned}

gives $\mu, \sigma^2 | y$ which is what we want.

Move to MCMC. From 2013 `rstan` from Columbia gives a quantum leap for semiparametric models over BUGS etc.

## Lab

```{r}
pTrue <- 0.37
set.seed(1)
x <- rbinom(100, 1, pTrue)
print(x)

library(rstan)
rstan_options(auto_write = TRUE)
Sys.setenv(LOCAL_CPPFLAGS = '-march=native')
options(mc.cores = parallel::detectCores())

# compile code
myf <- file.path("lab3a.stan")
mod <- rstan::stan_model(myf ,verbose = F)

fit <- rstan::sampling(object  = mod,
                       data    = list(x = x),
                       warmup  = 1000,
                       chains  = 1,
                       thin    = 1,
                       iter    = 2000,
                       refresh = 100)

pMCMC <- extract(fit, "p", permuted = F)
plot(pMCMC, type = "l", col = "darkgreen")
plot(pMCMC[1:999], pMCMC[2:1000], col = "blue")
acf(pMCMC)
monitor(pMCMC)
hist(pMCMC, col = "palegreen", breaks = 50, prob = T)

```


### Beta binomial conjugate approach

We can do the above just using a conjugate prior to the binomial likeli.

```{r}
xg <- seq(0, 1, length = 10001)
yg <- dbeta(xg, sum(x) + 1, 101 - sum(x))
lines(xg, yg, col = "blue", lwd = 3)
abline(v = pTrue, col = "red", lwd = 3)
```



```{r}
library(HRW)
summMCMC(list(pMCMC), parNames = list("p"))

```


### Back to Warsaw apt

```{r}
library(HRW)  ;  
library(rstan) 

rstan_options(auto_write = TRUE)
Sys.setenv(LOCAL_CPPFLAGS = '-march=native')
options(mc.cores = parallel::detectCores())

# Load Warsaw apartment data:

data(WarsawApts)

# Standardize both predictor and response variable
# and set hyperparameter values:

xOrig <- WarsawApts$construction.date
yOrig <- WarsawApts$areaPerMzloty
mean.x <- mean(xOrig)  ; sd.x <- sd(xOrig)
mean.y <- mean(yOrig)  ; sd.y <- sd(yOrig)
x <- (xOrig - mean.x)/sd.x
y <- (yOrig - mean.y)/sd.y
sigmaBeta <- 1e5 ; Au <- 1e5 ; Aeps <- 1e5

# Obtain linear and spline basis design matrices (X and Z):

X <- cbind(1,x)
aOrig <- min(xOrig) ; bOrig <- max(xOrig)
a <- (aOrig - mean.x)/sd.x  ;  b <- (bOrig - mean.x)/sd.x
numIntKnots <- 25
intKnots <-  quantile(unique(x),seq(0,1,length=numIntKnots+2)
                      [-c(1,numIntKnots+2)])
Z <- ZOSull(x,intKnots=intKnots,range.x=c(a,b))
ncZ <- ncol(Z)

```



Compile model

```{r}
allData <- list(n=length(x),
                ncZ=ncZ,
                y=y,
                X=X,
                Z=Z,
                sigmaBeta=as.numeric(sigmaBeta),
                Au=Au,
                Aeps=Aeps)

myf <- file.path("lab3b.stan")
mod <- rstan::stan_model(myf ,verbose = F)

fit <- rstan::sampling(object  = mod,
                       data    = allData,
                       warmup  = 1000,
                       chains  = 1,
                       thin    = 1,
                       iter    = 3000,
                       refresh = 100)

summary(fit)$summary

```



```{r}
betaMCMC <- NULL
for (j in 1:2)
{
   charVar <- paste("beta[",as.character(j),"]",sep="") 
   betaMCMC <- rbind(betaMCMC,extract(fit,charVar,permuted=FALSE))
}
uMCMC <- NULL
for (k in 1:ncZ)
{
   charVar <- paste("u[",as.character(k),"]",sep="") 
   uMCMC <- rbind(uMCMC,extract(fit,charVar,permuted=FALSE))
}
sigmaepsMCMC <- as.vector(extract(fit,"sigmaeps",permuted=FALSE))
sigmauMCMC <- as.vector(extract(fit,"sigmau",permuted=FALSE))

# Obtain MCMC samples of regression curves over a fine grid:

ng <- 101
xgOrig <- seq(aOrig,bOrig,length=ng)
xg <- (xgOrig - mean.x)/sd.x
Xg <- cbind(rep(1,ng),xg)
Zg <- ZOSull(xg,intKnots=intKnots,range.x=c(a,b))
fhatMCMC <- Xg%*%betaMCMC + Zg%*%uMCMC

# Convert fhatMCMC matrix to original scale:

fhatMCMCOrig <- fhatMCMC*sd.y + mean.y
fhatgOrig <- apply(fhatMCMCOrig,1,mean)
credLower <- apply(fhatMCMCOrig,1,quantile,0.025)
credUpper <- apply(fhatMCMCOrig,1,quantile,0.975)

# Display the fit:

par(mai=c(1,1.1,0.1,0.1))
cex.labVal <- 2   ;   cex.axisVal <- 1.5
plot(xOrig,yOrig,type="n",xlab="construction date (year)",
     ylab="area (square meters) per million zloty",
     bty="l",xlim=range(xgOrig),ylim=range(c(credLower,credUpper,yOrig)),
     cex.lab=cex.labVal,cex.axis=cex.axisVal)
polygon(c(xgOrig,rev(xgOrig)),c(credLower,rev(credUpper)),
        col="palegreen",border=FALSE)
lines(xgOrig,fhatgOrig,col="darkgreen",lwd=2)
points(xOrig,yOrig,col="dodgerblue")
abline(v=quantile(xOrig,0.25),lty=2,col="darkorange")
abline(v=quantile(xOrig,0.50),lty=2,col="darkorange")
abline(v=quantile(xOrig,0.75),lty=2,col="darkorange")

# cat("Hit Enter to continue.\n")
# ans <- readline()

# Obtain samples from the posterior distribution of the
# effective degrees of freedom:

X <- cbind(rep(1,length(x)),x)
Z <- ZOSull(x,intKnots=intKnots,range.x=c(a,b))
CTC <- crossprod(cbind(X,Z)) ; Dmat <- diag(c(0,0,rep(1,ncol(Z))))
lambdaMCMC <- (sigmaepsMCMC/sigmauMCMC)^2
EDFMCMC <- rep(NA,length(lambdaMCMC))
for (i in 1:length(lambdaMCMC))
   EDFMCMC[i] <- sum(diag(solve(CTC+lambdaMCMC[i]*Dmat,CTC)))

# Convert error standard deviation MCMC sample to the orginal units:

sigmaepsOrigMCMC <- sd.y*sigmaepsMCMC

# Do some summaries and diagnostic checking of the MCMC:

indQ1 <- length(xgOrig[xgOrig<quantile(xOrig,0.25)])
indQ2 <- length(xgOrig[xgOrig<quantile(xOrig,0.50)])
indQ3 <- length(xgOrig[xgOrig<quantile(xOrig,0.75)])
fhatOrigQ1MCMC <- fhatMCMCOrig[indQ1,]
fhatOrigQ2MCMC <- fhatMCMCOrig[indQ2,]
fhatOrigQ3MCMC <- fhatMCMCOrig[indQ3,]
MCMClist <- list(cbind(EDFMCMC,sigmaepsOrigMCMC,
                 fhatOrigQ1MCMC,fhatOrigQ2MCMC,fhatOrigQ3MCMC))
parNamesVal <- list(c("effective","degrees","of freedom"),
                    c("error","standard","deviation"),
                    c("reg'n func. est.","at 1st quartile","of construc. date"),
                    c("reg'n func. est.","at 2nd quartile","of construc. date"),
                    c("reg'n func. est.","at 3rd quartile","of construc. date"))

summMCMC(MCMClist,parNames=parNamesVal)



# Obtain chain summaries via the monitor() function:

myMCMCarray <- array(0,dim=c(length(sigmaepsMCMC),1,5))
myMCMCarray[,1,1] <- EDFMCMC
myMCMCarray[,1,2] <- sigmaepsOrigMCMC
myMCMCarray[,1,3] <- fhatOrigQ1MCMC
myMCMCarray[,1,4] <- fhatOrigQ2MCMC
myMCMCarray[,1,5] <- fhatOrigQ3MCMC
monitorAnswer <- monitor(myMCMCarray,warmup=0,print=FALSE)
dimnames(monitorAnswer)[[1]] <- c("EDF","err. st. dev.","f(Q_1)","f(Q_2)","f(Q_3)")
print(signif(monitorAnswer,4))
```


