---
title: "Semiparametric Regression"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Lab 4

GAM are a method that allows the data to choose the marginal based on splines. The idea is to apply a penalized spline to each covariate. The only assumption (here) is additivity; we do not look at interactions yet. This makes for easier interpretation.

## Lab

Plots show the PS fits with generalised cross-val used to choose the amount of penalisation for each fit.

```{r}
# install.packages(c("HRW","Ecdat","mlbench","quantreg",
#                    "gam","gamlss","gamsel"),dependencies=TRUE)
library(HRW)
library(Ecdat)
data(Caschool)

Caschool$log.avginc <- log(Caschool$avginc)
pairs(Caschool[, c(17, 7, 8, 11, 12, 13, 
                   15, 18)], col = "dodgerblue")
library(mgcv)

fitGAM <- gam(mathscr ~ s(calwpct) + 
                s(mealpct) + s(compstu) + s(expnstu) + 
                s(str) + s(elpct) +
                s(log.avginc), data = Caschool)

plot(fitGAM, pages = 1, shade = T, 
     col = "darkgreen", shade.col = "palegreen")

print(summary(fitGAM))
```

The summary gives the effective df of each additive model component. The linear effects have edf equal to one. The non-linear effects have higher edf values. Only a few of the terms are significant.

### Model selection

Stepwise. First specify that we want to look at terms not being in the model, being in the model linearly and being in the model non-lin with 2 edf. AIC iss used to id model.

```{r}
scopeList <- list(
  "calwpct" =    ~ 1 + calwpct + s(calwpct,2),
  "mealpct" =    ~ 1 + mealpct + s(mealpct,2),
  "compstu" =    ~ 1 + compstu + s(compstu,2),
  "expnstu" =    ~ 1 + expnstu + s(expnstu,2),
  "str" =        ~ 1 + str + s(str,2),
  "log.avginc" = ~ 1 + log.avginc + s(log.avginc,2),
  "elpct" =      ~ 1 + elpct + s(elpct,2))
```

```{r}
detach("package:mgcv")
library(gam)
fitInit <- gam(mathscr ~ calwpct + 
                mealpct + compstu + expnstu + 
                str + elpct +
                log.avginc, data = Caschool)
stepFit <- step.Gam(fitInit, scope = scopeList)

names(stepFit$"model")[-1]

```

```{r}
detach("package:gam")
library(mgcv)
fitStepGam <- gam(mathscr ~ mealpct  + elpct +
                 s(calwpct, k = 27) +
                 s(compstu, k = 27) +
                 s(log.avginc, k = 27) , 
               data = Caschool)
plot(fitStepGam, pages = 1)
```

## Mortgage applications


```{r}
library(HRW)
data("BostonMortgages")

d <- BostonMortgages


pdeny_blk <- sum(d$deny[d$black == "yes"] == "yes") / sum(d$black == "yes")

fitSimpGLM <- glm(deny ~ black,family = binomial,
                  data = d)
betaHat <- fitSimpGLM$coef[2]
SEbetaHat <- summary(fitSimpGLM)$coefficients[2,2]
CIandEstForLogOR <- c(betaHat - 
                    qnorm(0.975)*SEbetaHat,
                    betaHat,
                    betaHat + 
                    qnorm(0.975)*SEbetaHat)
CIandEstForOR <- as.numeric(exp(CIandEstForLogOR))
names(CIandEstForOR) <- c("lower","estimate",
                          "upper")
print(round(CIandEstForOR,2))
```

The above gives the odds ratio of black versus nun-black mortgage applicants being unsuccessful, which equals 4. Next more variables are incorporated.

Use `step.Gam()` from gam package to select a model allowing predictors:
+ dir
+ hir
+ lvr
with 2, 6 or 10 edf.

```{r}
library(HRW)
data(BostonMortgages)
detach("package:mgcv") ; 
library(gam)

fitInitial <- gam(deny ~ dir + hir + lvr 
                   + factor(ccs) + factor(mcs) 
                   + factor(pbcr)
                   + dmi + self 
                   + single + uria
                   + condominium + black,
                   family = binomial,
                   data = BostonMortgages)

scopeList <- list(
        "dir" = ~ 1+dir+s(dir,2)+s(dir,6)+s(dir,10),
        "hir" = ~ 1+hir+s(hir,2)+s(hir,6)+s(hir,10),
        "lvr" = ~ 1+lvr+s(lvr,2)+s(lvr,6)+s(lvr,10),
        "ccs" = ~ 1 + factor(ccs),
        "mcs" = ~ 1 + factor(mcs),
        "pbcr" = ~ 1 + factor(pbcr),
        "dmi" =  ~ 1 + dmi,
        "self" = ~ 1 + self,
        "single" = ~ 1 + single,
        "uria" =   ~ 1 + uria,
        "condominium" =  ~ 1 + condominium,
        "black" =  ~ 1 + black)

stepFit <- step.Gam(fitInitial,scope = scopeList)

print(names(stepFit$"model")[-1])

```

Nine parameters are chosen, two of which have high non-linearity.


```{r}
library(HRW)
data(BostonMortgages)
detach("package:gam") ; 
library(mgcv)

fitGAM <- gam(deny ~ s(dir, k = 27) + 
                s(lvr, k = 27) +
                factor(ccs) + 
                factor(pbcr) +
                dmi + self + single + uria + black,
              family = binomial,
              data = BostonMortgages)
print(summary(fitGAM))
     
options(warn = -1)
plot(fitGAM, pages = 1, xlim = c(0, 1), shade = T, 
     shade.col = "palegreen", col = "darkgreen")
```

The above figure just shows the non-linear parts (dir and lvr). Now extract the ORs.


```{r}
betaHat <- fitGAM$coefficients[12]
SEbetaHat <- summary(fitGAM)$p.table[12, 2]
CIandEstForLogOR <- c(betaHat - qnorm(0.975) * SEbetaHat, 
                      betaHat, 
                      betaHat + qnorm(0.975) * SEbetaHat)
CIandEstForOR <- as.numeric(exp(CIandEstForLogOR))
names(CIandEstForOR) <- c("lwr", "est", "upr")

print(round(CIandEstForOR, 2))
```

So now the OR for black has dropped to be about half of what it was previously. However, the OR is still > 1.

To get a plot on the probability scale do:

```{r}
demo(BostMortGAMfit, package = "HRW")

system.file("demo", "BostMortGAMfit.R", package = "HRW")
```


## Count regression

```{r}
library(Ecdat)
data(OFP)

# Take ofp (num of physician office visits) as the response
par(mfrow = c(1, 1))
hist(OFP$ofp, col = "blue", breaks = 100)
mean(OFP$ofp)
var(OFP$ofp)

```

In `mgcv`, overdispersion can be handled with `family = quasipoisson` another al is to use neg bin rather than poisson but neither mgcv nor gam support nb. Not all is lost because gamlss has `family = NBI`. Here is example.

```{r}
OFPforAna <- OFP
OFPforAna$age <- 10 * OFPforAna$age
OFPforAna <- OFPforAna[OFPforAna$age < 95, ]
library(gamlss)
fitgamlss <- gamlss(ofp ~ pb(age) + pb(school) +
                      black + sex + maried + 
                      adldiff + privins + medicaid +
                      region + hlth, family = NBI, 
                    data = OFPforAna)
summary(fitgamlss)
```

Most of the terms are sig. The plot function shows the residuals but does not give an easy way to visualise the relationship between the response and the terms in the model.


## Model selection via gamsel

Instead of stepwise on AIC use lasso for regularisation.

```{r}
library(Ecdat)
data("Caschool")
Caschool$log.avginc <- log(Caschool$avginc)
X <- Caschool[, c(7,8,11,12,13, 15, 18)]
y <- as.vector(Caschool[, 17])
library(gamsel)
set.seed(919)
fitCVgamsel <- cv.gamsel(X, y)
fitCVgamsel$gamsel.fit

plot(fitCVgamsel)
```

The plot shows the cv mse with +/- 1 se as a func of $\lambda$. Lambda min is equal to around 15 and 1se is 47. The latter takes the variability of the cv mse function into account so we choose the larger value.

```{r}
fitCVgamsel$lambda.min
fitCVgamsel$lambda.1se
fitCVgamsel$index.1se
```

```{r}
par(mfrow=c(2,4)); indexVal <- fitCVgamsel$index.1se
for (iPred in 1:7)
   plot(fitCVgamsel$gamsel.fit,newx = X,
        which = iPred,index = fitCVgamsel$index.1se,
        col.main = "navy",main = names(X)[iPred],
        rugplot = TRUE,bty = "l",
        ylim = c(-8,16))

```

The colour ccoding in the plot has blue indicating a zero effect, green a linear one and red a non-linear one.

## Model plots for gamlss

Note that variability bands cannot be obtained because gamlss predict function does not provide them. 

```{r}
OFPforAna <- OFP
OFPforAna$age <- 10 * OFPforAna$age
OFPforAna <- OFPforAna[OFPforAna$age < 95, ]
library(gamlss)
fitgamlss <- gamlss(ofp ~ pb(age) + pb(school) +
                      black + sex + maried + 
                      adldiff + privins + medicaid +
                      region + hlth, family = NBI, 
                    data = OFPforAna)
summary(fitgamlss)
```

```{r}
modalValue <- function(x)
   return(unique(x)[which.max(tabulate(
          match(x,unique(x))))])
ng <- 101
ageg <- seq(min(OFPforAna$age),max(OFPforAna$age),
            length = ng)
schoolg <- seq(min(OFPforAna$school),
               max(OFPforAna$school),length = ng)
factorsAtModes <- data.frame(
         black  = modalValue(OFPforAna$black),
         sex  = modalValue(OFPforAna$sex),
         maried  = modalValue(OFPforAna$maried),
         adldiff = modalValue(OFPforAna$adldiff),
         privins  = modalValue(OFPforAna$privins),
         medicaid  = modalValue(OFPforAna$medicaid),
         region  = modalValue(OFPforAna$region),
         hlth  = modalValue(OFPforAna$hlth))
newdataAge <- cbind(age = ageg,
                   school = mean(OFPforAna$school),
                   factorsAtModes)
fageg <- predict(fitgamlss,newdata = newdataAge,
                 type = "response")
newdataSchool <- cbind(age = mean(OFPforAna$age),
                 school = schoolg,factorsAtModes)
fschoolg <- predict(fitgamlss,
            newdata = newdataSchool,
            type = "response")
dev.off() ; par(mfrow = c(1,2))
plot(ageg,fageg,type = "l",bty = "l",
     col = "darkgreen",lwd =2,
     xlab = "age",
     ylab = "mean no. of physician office visits",
     ylim = c(3,8))
rug(jitter(OFPforAna$age),col = "dodgerblue")
plot(schoolg,fschoolg,type = "l",bty = "l",
     col = "darkgreen",lwd =2,
     xlab = "number of years of education",
     ylab = "mean no. of physician office visits",
     ylim = c(3,8))
rug(jitter(OFPforAna$school),col = "dodgerblue")

```

The LHS panel shows the effect of age on visits. Num of visits increase as a function of num years edu, which may be treated as a proxy for ses.


