---
title: "Semiparametric Regression"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Lab 4

GAM are a method that allows the data to choose the marginal based on splines. The idea is to apply a penalized spline to each covariate. The only assumption (here) is additivity; we do not look at interactions yet. This makes for easier interpretation.

## Lab

Plots show the PS fits with generalised cross-val used to choose the amount of penalisation for each fit.

```{r}
# install.packages(c("HRW","Ecdat","mlbench","quantreg",
#                    "gam","gamlss","gamsel"),dependencies=TRUE)
library(HRW)
library(Ecdat)
data(Caschool)

Caschool$log.avginc <- log(Caschool$avginc)
pairs(Caschool[, c(17, 7, 8, 11, 12, 13, 
                   15, 18)], col = "dodgerblue")
library(mgcv)

fitGAM <- gam(mathscr ~ s(calwpct) + 
                s(mealpct) + s(compstu) + s(expnstu) + 
                s(str) + s(elpct) +
                s(log.avginc), data = Caschool)

plot(fitGAM, pages = 1, shade = T, 
     col = "darkgreen", shade.col = "palegreen")

print(summary(fitGAM))
```

The summary gives the effective df of each additive model component. The linear effects have edf equal to one. The non-linear effects have higher edf values. Only a few of the terms are significant.

### Model selection

Stepwise. First specify that we want to look at terms not being in the model, being in the model linearly and being in the model non-lin with 2 edf. AIC iss used to id model.

```{r}
scopeList <- list(
  "calwpct" =    ~ 1 + calwpct + s(calwpct,2),
  "mealpct" =    ~ 1 + mealpct + s(mealpct,2),
  "compstu" =    ~ 1 + compstu + s(compstu,2),
  "expnstu" =    ~ 1 + expnstu + s(expnstu,2),
  "str" =        ~ 1 + str + s(str,2),
  "log.avginc" = ~ 1 + log.avginc + s(log.avginc,2),
  "elpct" =      ~ 1 + elpct + s(elpct,2))
```

```{r}
detach("package:mgcv")
library(gam)
fitInit <- gam(mathscr ~ calwpct + 
                mealpct + compstu + expnstu + 
                str + elpct +
                log.avginc, data = Caschool)
stepFit <- step.Gam(fitInit, scope = scopeList)

names(stepFit$"model")[-1]

```

```{r}
detach("package:gam")
library(mgcv)
fitStepGam <- gam(mathscr ~ mealpct  + elpct +
                 s(calwpct, k = 27) +
                 s(compstu, k = 27) +
                 s(log.avginc, k = 27) , 
               data = Caschool)
plot(fitStepGam, pages = 1)
```

## Mortgage applications


```{r}
library(HRW)
data("BostonMortgages")



```

