---
title: "Semiparametric Regression"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Lab 5

Example is bone mineral data that has a few points for each individual. Assume we have $m$ groups with $n_i$ in each group.

\begin{aligned}
SBMD_{ij} = U_i + f(age_{ij}) + e_{ij}
\end{aligned}

The Z matrix now has two parts, the random intercepts and the spline basis functions.

Other example is adolescent growth curves. Here we have a lot more measures per person, but they are not parallel so we need a bit more than a random intercept.

\begin{aligned}
y_{ij} = g_i(x_{ij}) + f(age_{ij}) + e_{ij}
\end{aligned}

$f$ is the population curve. $g_i$ are the deviations. These are called group specific growth curve models.

But these cannot be fit using `mgcv`, `gam` and `gamlss` so we need to use nlme or stan. For nlme we have to set up the matrices ourselves.


```{r}
library(lattice) ; library(HRW); data(femSBMD)
pobj <- xyplot(spnbmd ~ age|factor(ethnicity),
        group = idnum, data = femSBMD,
        xlab = "age",
        ylab = "spinal bone mineral density",
        panel = function(x,y,subscripts,groups)
        {
           panel.grid()
           panel.superpose(x,y,subscripts,groups,
           type = "b",pch = 16,lwd = 2)
        })
print(pobj)
```

The idnum term contains the group level ids. Let's say we are intersted in the differences in mean spinal bone mineral density among the 4 ethnic categories after accounting for age and within group correl.

```{r}
library(mgcv)
fitGAMM <- gamm(spnbmd ~ s(age) + black + 
                  hispanic + white,
                random = list(idnum = ~1), 
                data = femSBMD)
plot(fitGAMM$gam, shade  = T, shade.col = "palegreen")
```

Note the vertical centering. A summary can be obtained with:

```{r}
summary(fitGAMM$gam)
intervals(fitGAMM$lme)
```

So, the comparison between asian and black looks significant, but there is no ev for difference between asian and hisp nor asian and white.

### Group specific curves

Useful ref Donnelly 1995 Prediction and creation of smooth curves for temporally correlated longitudinal data but here we follow that of Durban 2005 Simple fitting of sub-specific curves for longitudinal data.

```{r}
library(HRW)  ;  library(lattice)
data(growthIndiana)
growthINblackMales <- 
   growthIndiana[(growthIndiana$male == 1)
        & (growthIndiana$black == 1),]
pobj <- xyplot(height ~ age|idnum,groups = idnum,
        data = growthINblackMales,
        layout = c(4,7),strip = FALSE,
        xlab = "age",
        ylab = "height (centimetres)",
        as.table = TRUE,
        panel = function(x,y,subscripts,groups)
        {
           panel.grid()
           panel.superpose(x,y,subscripts,groups,
                col = "dodgerblue",type = "b")
        })
print(pobj)
```


Note that the curves differ markedly.

The group specific models can proceed from:


\begin{aligned}
y_{ij} = \beta_0 + \beta_1 x 
\end{aligned}


<!-- y_{g, k} \sim N(0, \sigma_g^2) \\ -->
<!-- g_i(x_i) = U_{0i} + U_{1i}x +  \sum_{k=1}^K_{r}u_{r, ik} z_{r, k}(x) \\ -->
<!-- U \sim N(0, \Sigma) \\ -->
<!-- u_{r, ik} \sim N(0, \sigma_r^2) -->

We use `lme()` to fit the above. 

```{r}
library(nlme)

age <- growthINblackMales$age 
height <- growthINblackMales$height
idnumOrig <- growthINblackMales$idnum
idnum <- match(idnumOrig, unique(idnumOrig))
growthINblackMales$idnum <- idnum
# set up the z_gbl and the z_grp

numObs <- length(height)
numIntKnotsGbl <- 20
intKnotsGbl <- quantile(unique(age),
               seq(0,1,length = numIntKnotsGbl + 2)
               [-c(1,numIntKnotsGbl + 2)])
range.age <- c(5.5,20)
Zgbl <- ZOSull(age,range.x = range.age,
               intKnots = intKnotsGbl)
numIntKnotsGrp <- 10	
intKnotsGrp <- quantile(unique(age),
   seq(0,1,length = numIntKnotsGrp + 2)
   [-c(1,numIntKnotsGrp+2)])
Zgrp <- ZOSull(age,range.x = range.age,
        intKnots = intKnotsGrp)

# set up the random effect structure

dummyID <- factor(rep(1, numObs))
Zblock <- list(dummyID = pdIdent(~ -1 + Zgbl),
               idnum = pdSymm(~ age),
               idnum = pdIdent(~ -1 + Zgrp))

```

The dummyID is an array of len numObs (total num of obs) with all entries equal to 1. This tricks lme() into accomodating a global pen spline component.

The `pdIdent(~ -1 + Zgbl)` invokes the multiple of I $u_{gbl} \sim N(0, \sigma^2_{gbl}I)$ across the entire data regardless of within sub grouping.

The `pdSymm(~ age)` invokes the block diag unstruc 2x2 cov matrix on [U_{0i} U_{1i}]^T

The `pdIdent(~ -1 + Zgrp)` does the same for the group level u.

Now call lme:

```{r}
blkMalGD <- groupedData(height ~ age | dummyID, 
                        data = data.frame(height, age, 
                                          Zgbl, Zgrp, idnum))
fitGSCM <- lme(height ~ age, data = blkMalGD, random = Zblock)
```

Now we can plot the subject level curves.

```{r}
ng <- 101
ageg <- seq(range.age[1], range.age[2], length = ng)
Xg <- cbind(1, ageg)
Zgblg <- ZOSull(ageg, range.x = range.age, intKnots = intKnotsGbl)
Zgrpg <- ZOSull(ageg, range.x = range.age, intKnots = intKnotsGrp)

# extract coef

betaHat <- as.vector(fitGSCM$coefficients$fixed)
uHat <- as.vector(fitGSCM$coefficients$random[[1]])
fHatg <- as.vector(Xg %*% betaHat + Zgblg %*% uHat)

```

Code for plotting.

```{r}
numGrp <- length(unique(idnum))
curvEsts <- vector("list",numGrp)
for (i in 1:numGrp)
{
   uLinHati <- as.vector(
               fitGSCM$coef$random[[2]][i,])
   uSplHati <- as.vector(
               fitGSCM$coef$random[[3]][i,])
   gHati <- as.vector(Xg%*%uLinHati 
                    + Zgrpg%*%uSplHati)
   curvEsts[[i]] <- fHatg + gHati
}

pobj <- xyplot(height ~ age|idnum,
        groups = idnum,
        data = growthINblackMales,
        strip = FALSE,
        xlab = "age (years)",
        ylab = "height (centimetres)",
        as.table = TRUE, layout = c(4,7),
        panel = function(x,y,subscripts,groups)
        {  
           panel.grid()
           adolNum <- idnum[subscripts][1]
           panel.superpose(x,y,subscripts,groups,
           col = "dodgerblue",type = "b")
           panel.xyplot(ageg,curvEsts[[adolNum]],
           col = "darkgreen",lwd = 2,type = "l")
        })
print(pobj)

```

We can check the residuals with:

```{r}
plot(fitGSCM)
```

### Contrast Curves

Again using the indiana growth data. This is purely demo because there is too much code to type a custom version. The code stratifies by sex.

```{r}
library(HRW)
demo(maleGrowthIndiananlme, package = "HRW")
```

The code can be obtained from the following file.

```{r}
system.file("demo", "maleGrowthIndiananlme.R", package = "HRW")
```

A Bayesian version can be found here:

```{r}
system.file("demo", "maleGrowthIndianaBayes.R", package = "HRW")
```

Finally, a stan version can be found here but takes hours to run.

```{r, eval=F}
library(HRW)
system.file("demo", "coralAna.R", package = "HRW")
```

