---
title: "Semiparametric Regression"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Lab 6

Now we look at bivariate setting.

```{r}
library(HRW)
library(lattice)
data("ozoneSub")

x <- ozoneSub$longitude
y <- ozoneSub$latitude
z <- ozoneSub$ozone

cloud(z ~ x*y, xlab = "deg long",
      ylab = "deg lat",
      zlab = "ozone")
```

Fit model

```{r}
library(mgcv)
fit <- gam(z ~ s(x, y))
plot(fit)
points(x, y, col = "dodgerblue")
gam.check(fit)
```


The residual plot looks ok but the low p-value for k-index is worrying and indicates insufficient spline basis functions.

```{r}
fitHighBasis <- gam(z ~ s(x, y, k = 60))
summary(fitHighBasis)
plot(fitHighBasis)
points(x, y, col = "dodgerblue")
gam.check(fitHighBasis)
```

Now contour plot.

```{r}
ngrid <- 201
xgrid <- seq(min(x),max(x),length = ngrid)
ygrid <- seq(min(y),max(y),length = ngrid)
xymesh <- expand.grid(xgrid,ygrid)
names(xymesh) <- c("x","y")
fitmesh <- matrix(predict(fitHighBasis,
                  newdata = xymesh),
                  ngrid,ngrid)
filled.contour(xgrid,ygrid,fitmesh,
               xlab = "degrees longitude",ylab = "degrees latitude",
               key.title = title(main = "ozone"))
```

###  Advanced Graphical displays

Most geo datasets are dispersed.

```{r}
library(HRW)
data("SydneyRealEstate")
x <- SydneyRealEstate$longitude
y <- SydneyRealEstate$latitude
z <- SydneyRealEstate$logSalePrice

plot(x, y, col = "dodgerblue", cex = 0.1)


data("SydneyRealEstateBdry")
print(SydneyRealEstateBdry)
lines(SydneyRealEstateBdry)
```

Fit model.

```{r}
library(mgcv)
fit <- gam(z ~ s(x, y, k = 150))
plot(fit)
```

Image plot of surface:

```{r}
ngrid <- 201
xgrid <- seq(min(x), max(x), length = ngrid)
ygrid <- seq(min(y), max(y), length = ngrid)
xymesh <- expand.grid(xgrid, ygrid)
names(xymesh) <- c("x", "y")
fitmesh <- matrix(predict(fit, newdata = xymesh), ngrid, ngrid)
image(xgrid, ygrid, fitmesh, 
      col = terrain.colors(1000),
      xlab = "deg long",
      ylab = "deg lat")
points(x, y, col = "dodgerblue", pch = ".")



```

Clip image with boundary.

```{r}
outInds <- (1:ngrid^2)[pointsInPoly(xymesh, 
                                    SydneyRealEstateBdry)==F]
fitmesh[outInds] <- NA
image(xgrid, ygrid, fitmesh, 
      col = terrain.colors(1000),
      xlab = "deg long",
      ylab = "deg lat")

lines(SydneyRealEstateBdry)
```

### Creating polygon regions for clipping

Plot shows the long/lat of data on scallop abundance in US.

```{r}
library(HRW)
data("scallop")
x <- scallop$longitude
y <- scallop$latitude

plot(x, y, col = "dodgerblue", cex = 0.1)

```

Create boundary. The bit commented out is interactive - you need to pick the points for the boundary.

```{r}
# myScallopBdry <- createBoundary(x, y)
# print(myScallopBdry)
# write.table(myScallopBdry, "bdry.txt", col.names = F, row.names = F)
mys <- read.table("bdry.txt")
plot(x, y, col = "dodgerblue")
lines(mys, col = "navy")
```


